---
title: "Class 3 Example regarding 2017 Survey Data"
author: "Thomas E. Love"
date: "Generated `r Sys.Date()`"
output:
  html_document:
    toc: yes
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(comment=NA)
options(width = 70)

library(magrittr); library(tidyverse)
## make sure these libraries are installed in R
```

```{r load_data}
sur_day1 <- read.csv("surveyday1_2017.csv") %>% tbl_df()

sur_day1
```

# The variables we're studying today

```{r list_of_variables_with_summary}
sur_day1 %>%
    select(year, sex, favcolor, haircut, ageguess, pulse, lastsleep) %>%
    summary()
```

So we have missing values in the `favcolor`, `haircut`, `ageguess`, `pulse` and `lastsleep` data to watch out for.

# What are the favorite colors of students in 431?

First, we'll try the `summary` function.

```{r favcolors_summaries}
sur_day1 %>%
    select(favcolor) %>%
    summary()
```

Perhaps a `table` would be more useful here, since the data are categorical.

```{r favcolors_table}
sur_day1 %>%
    select(favcolor) %>%
    table()
```

It might be nice to see that table just for the 2017 data.

```{r favcolors_table_2017_only}
sur_day1 %>%
    filter(year == 2017) %>%
    select(favcolor) %>%
    table()
```

Let's drop the levels with zeros, and add a total.

```{r favcolors_table_2017_improved}
sur_day1 %>%
    filter(year == 2017) %>%
    select(favcolor) %>%
    droplevels() %>%
    table() %>%
    addmargins()
```

Another approach that creates a tibble with the counts, so that we can arrange them in descending order from most common to least common?

```{r favcolors_table_2017_alternative}
sur_day1 %>%
    filter(year == 2017) %>%
    group_by(favcolor) %>%
    count() %>%
    arrange(desc(n))
```

## A more challenging question

What is the percentage of students who prefer `blue` in each year?

```{r blue_favcolors_by_year}
sur_day1 %>%
    group_by(year) %>%
    summarize(blue_count = sum(favcolor == "blue", na.rm=TRUE), 
              total_count = sum(!is.na(favcolor)), 
              blue_percent = round(100 * blue_count / total_count, 1)) 
```

# What price did students in 431 pay for a haircut?

Let's summarize this graphically first. Perhaps a histogram?

```{r histogram_haircut_draft}
ggplot(data = sur_day1, aes(x = haircut)) +
    geom_histogram()
```

## Improving the Histogram

Let's change the fill and color, and specify the size of the bins to be $10.

```{r histogram_haircut_revision_A}
ggplot(data = sur_day1, aes(x = haircut)) +
    geom_histogram(binwidth = 10, 
                   fill = "blue", color = "white")
```

Can we improve the labels?

```{r histogram_haircut_revision_B}
ggplot(data = sur_day1, aes(x = haircut)) +
    geom_histogram(binwidth = 10, 
                   fill = "blue", color = "white") +
    labs(x = "Price of Last Haircut in $", 
         y = "Number of Students")
```

Can we eliminate that missing values issue?

```{r histogram_haircut_revision_C}
sur_day1 %>%
    filter(!is.na(haircut)) %>%
ggplot(., aes(x = haircut)) +
    geom_histogram(binwidth = 10, fill = "blue", color = "white") +
    labs(x = "Price of Last Haircut in $", y = "Number of Students")
```

## Numerical Summary

```{r haircut_numerical_summary}
sur_day1 %>%
    select(haircut) %>%
    summary()
```

An alternative approach, using the `favstats` function from the `mosaic` package, and the `%$%` pipe rather than the usual `$>$`:

```{r haircut_favstats}
sur_day1 %$% mosaic::favstats(haircut)
```

Or, you could use the `$` notation, as:

```{r haircut_favstats_alternative}
mosaic::favstats(sur_day1$haircut)
```


## A more challenging question: Haircut Prices by Year

Can we build a separate plot for each year?

```{r histogram_haircut_by_year}
sur_day1 %>%
    filter(!is.na(haircut)) %>%
    ggplot(., aes(x = haircut, fill = year)) +
        geom_histogram(binwidth = 10, color = "white") +
        guides(fill = FALSE) +
        labs(x = "Price of Last Haircut in $", y = "Number of Students") +
        facet_wrap(~ year)
```

Or we could use a boxplot instead, but this requires us to get R to see year as a factor...

```{r boxplot_haircut_by_year}
ggplot(sur_day1, aes(y = haircut, x = factor(year))) +
    geom_boxplot()
```

Numerical Summary by Year?

```{r haircut_summary_by_year}
sur_day1 %>% 
    filter(!is.na(haircut)) %>%
    group_by(year) %>%
    summarize(mean = mean(haircut), sd = sd(haircut), median = median(haircut))
```

# What is the relationship between sex and 431 students' guess of my age?

Here's a starting plot.

```{r scatterplot_ageguess_by_sex}
sur_day1 %>%
    filter(!is.na(ageguess)) %>%
ggplot(., aes(x = sex, y = ageguess)) +
    geom_point() 
```

Let's restrict to 2017, color the points, and jitter them left-to-right a bit.

```{r scatterplot_with_jitter_ageguess_by_sex}
sur_day1 %>%
    filter(!is.na(ageguess), year == "2017") %>%
ggplot(., aes(x = sex, y = ageguess, col = sex)) +
    geom_jitter(width = 0.2) 
```

A boxplot would be the usual summary of a quantitative variable's association with a categorical variable. We'll add a title, too.

```{r boxplot_age_guess_by_sex}
sur_day1 %>%
    filter(!is.na(ageguess), year == "2017") %>%
ggplot(., aes(x = sex, y = ageguess, fill = sex)) +
    geom_boxplot() +
    labs(title = "Guesses of Professor Love's Age, in 2017")
```

We might look at faceted histograms instead.

```{r histogram_age_guess_by_sex}
sur_day1 %>%
    filter(!is.na(ageguess), year == "2017") %>%
ggplot(., aes(x = ageguess, fill = sex)) +
    geom_histogram(binwidth = 3) +
    facet_wrap(~ sex) + 
    labs(title = "Guesses of Professor Love's Age, in 2017")
```

In the course notes, we'll also explore a joy plot in this setting.

# What is the relationship between 431 students' pulse rate and hours of sleep?

Here, we're looking at two quantitative variables. A scatterplot is usually the best choice.

```{r pulse_by_sleep}
ggplot(data = sur_day1, aes(x = pulse, y = lastsleep)) +
    geom_point()
```

Let's look at the 2017 data only, and fit a model to predict last sleep on the basis of pulse rate.

```{r pulse_by_sleep_for_2017_only_with_loess_model}
sur_day1 %>%
    filter(year == "2017") %>%
    ggplot(., aes(x = pulse, y = lastsleep)) +
    geom_point() +
    geom_smooth(method = "loess") +
    labs(title = "Hours of Sleep as a function of Pulse Rate",
         subtitle = "with fitted loess smooth")
```

We could instead restrict ourselves to a linear model for last sleep on the basis of pulse rate.

```{r pulse_by_sleep_for_2017_only_with_linear_model}
sur_day1 %>%
    filter(year == "2017") %>%
    ggplot(., aes(x = pulse, y = lastsleep)) +
    geom_point() +
    geom_smooth(method = "lm", col = "red") +
    labs(title = "Hours of Sleep as a function of Pulse Rate",
         subtitle = "with fitted linear model")
```

The correlation of `lastsleep` and `pulse` is likely to be of some interest. Note the use of both the `%>%` and `%$%` pipes in this case.

```{r correlation_lastsleep_pulse_in_2017}
sur_day1 %>%
    filter(year == "2017") %$%
    cor(pulse, lastsleep)
```

## Does the linear model change much by year?

```{r last_sleep_pulse_faceted_by_year}
sur_day1 %>%
    filter(!is.na(pulse), !is.na(lastsleep)) %>%
    ggplot(., aes(x = pulse, y = lastsleep, group = factor(year))) +
    geom_point() + 
    geom_smooth(method = "lm", color = "red") +
    facet_wrap(~ year)
```


